---
title: "Step-by-step discovery of Visium data analysis using embryonic heart sections."
author: Raphaël Mauron
date: today
format: html
execute:
  echo: true
editor_options:
  chunk_output_type: console

---

Intro, technology, polyA

## Set up

### Set project library

We start by setting up where the packages we need are etc...

```{r}
#| label: setup
#| include: false

.libPaths("/home/raphael.mauron/miniconda3/envs/VisiumHD/lib/R/library") # set project library

.libPaths() # check project library

lapply(.libPaths()[1], list.files) # check library
```


### Load library

```{r}
#| label: load packages

package.list <- c("semla", "Seurat", "viridis", "dplyr", "ggplot2", "gtools", "patchwork",
                  "RColorBrewer", "harmony", "tidyverse")
invisible(lapply(package.list, function(element) {
  library(element, character.only = TRUE)
}))
rm(package.list)
```


### Set seed for project

```{r}
set.seed(1)
```

## Loading my first Visium section

To analyse a Visium section, we need to load the data.
10x Genomics provides a softwares that takes the raw sequences and the microscopy images, 
and generates a count matrix (gene counts per bacrode), and several files that help us align 
the microscopy image to the counts data. 

Bellow we create a table that provides the path so R (and specific packages e.g., *semla*)
knows where these files are. We have here:
1. `filtered_feature_bc_matrix.h5` : the count data
2. `tissue_hires_image.png` : an H&E image that can be aligned to the counts
3. `tissue_positions_list.csv`: a table that contains the info to align the pixel of the image and the Visium spots
4. `scalefactors_json.json`: a ratio value that help us visualize the image and the spots in the right scale.

### Load data & create infoTable from metadata + spaceranger outs

These are all the paths we want:

```{r}
data_root_directory <- file.path("./data/spaceranger/V10F24-105_C1")

samples <- Sys.glob(paths = file.path(data_root_directory, "filtered_feature_bc_matrix.h5"))
imgs <- Sys.glob(paths = file.path(data_root_directory, "spatial", "tissue_hires_image.png"))
spotfiles <- Sys.glob(paths = file.path(data_root_directory, "spatial", "tissue_positions_list.csv"))
json <- Sys.glob(paths = file.path(data_root_directory, "spatial", "scalefactors_json.json"))
```


We create a table containing all these paths

```{r}
infoTable <- tibble(samples,
                    imgs,
                    spotfiles,
                    json)
```


We can now use `ReadVisiumData()` function from *semla* package to create an R-object with all the data we linked with their paths in the `infoTable`. *semla* uses the data structure from the well-known *Seurat* package, originally developed to facilitate the analysis of *single-cell RNA-seq* data. *semla* added, on top of this architecture, all the spatial information (image, coordinates, etc.) that are required for us to know where are the spots and to be able to later use this information in analyses and visualization.

Here we load the data and create an object.
We also load the image to the object.
And we will have a look at how things look:

```{r}
object <- ReadVisiumData(infoTable)
object <- LoadImages(object)
object
```


At this stage, we can already use and visualize the data, for instance plotting a gene:

```{r}
MapFeatures(
  object = object,
  layer = "counts",
  features = "MYH6", # <-- place to enter a gene name
  pt_size = 3,
  image_use = "raw"
)
```


### Question:

1. What is the dimension of the data?
2. How many spots does that section contain?
3. How many genes does that section contain?
4. Does it make sense?

`Hint`: look at the printed object!


### Cleaning up a bit...

We see that there are around the double unique genes compared to the number of human protein coding genes.
Since our data is coming from a poly-A capture technology, every transcripts in the sample 
that binds to a poly-T will be sequenced and counted. Among the extra things, there are some transcripts that we won't want to keep:

- Hemoglobin genes (raison)
- Mitochondrial genes (raison)
- ribosomal genes (raison)
- MALAT1 gene (raison)

In *R*, we can use simple `regular expression` (text patterns that match) to retrieve these genes:

```{r}
# find al the genes that correspond to mt, hb, rp and malat
undesirable_genes <- grep(pattern = c("^MT-|^HB|^RP|MALAT1"), x = rownames(object), value = T)

# remove two genes that correspond to our search but that we want to keep (growth factor & transcription factor)
undesirable_genes <- setdiff(undesirable_genes, c("HBEGF", "HBP1"))

# check the list of undesirable genes
undesirable_genes
```


We can get a list of all the other genes (that we want to keep), and see the difference compared to the original object.

```{r}
genes_to_keep <- setdiff(rownames(object), undesirable_genes)
length(rownames(object)) - length(genes_to_keep)
```


Now we can proceed with a first filtering. We want to remove the undesirable genes, meaning we want to keep only all the other genes.
We use the function `SubsetSTData()` from the *semla* package, which ensure that the spatial information is also modified correctly.

```{r}
object <- SubsetSTData(object, features = genes_to_keep)
object
```


### More systematic filtering

The extra transcripts we sequence and gather in our object do not limit itself to the 4 categories we succesfully removed above. Databases allow us to annotate the remaining genes we sequenced into some categories.
Here, we will use the  *GRCh38-2020-A genome reference* which let us figure out the main categories our biomolecules belongs too.

The *GRCh38-2020-A genome reference* is quite large so here we provide a smaller size of that reference that you can load in the next chunk. This chunk commented out is for you to see what has been done.

```{r}
#| eval: false

# package that help us load large files
library(rtracklayer)

# import the 
gtf <- import("/srv/home/10x_references/refdata-gex-GRCh38-2020-A/genes/genes.gtf")

unique(gtf$type)
# [1] gene           transcript     exon           CDS            start_codon   
# [6] stop_codon     UTR            Selenocysteine

annot <- gtf[gtf$type %in% c("gene")] |>
  mcols() |>
  as.data.frame() |>
  dplyr::select(gene_id, gene_name, gene_type) |>
  dplyr::distinct()

saveRDS(annot, "./data/GRCh38-2020-A_gene_biotypes.rds")
```


Here we load the lighter version of the reference genome:

```{r}
annot <- readRDS("./data/GRCh38-2020-A_gene_biotypes.rds")
head(annot)
unique(annot$gene_type)
```


Here we will use the information of the gene types to keep only the protein coding genes.
This step is project dependent, in other cases you might want to also keep the long-non-coding RNA or any types of immunoglobuline transcripts.

```{r}
# subset the reference genes to only the protein coding genes
annot_protein_coding <- subset(annot, gene_type == "protein_coding")

# fetch the gene names from our object that are annotated as protein coding genes in the reference genome
genes_to_keep_2 <- intersect(rownames(object), annot_protein_coding$gene_name)

# use the same subset idea as before
object <- SubsetSTData(object, features = genes_to_keep_2)
object
```


We can now see that our object significantly decrease. We can have a quick look at what genes are now present in our object:

```{r}
head(rownames(object), 20)
```


### Even more filtering

We now cleaned the genes that we are not interested in.
We can now analyse the distributions of these genes with the commonly used histograms, but also spatially.

```{r}
# check within the object more in details, we see the metadata
head(object@meta.data)
```

`nFeature_Spatial` = number of unique genes
`nCount_Spatial` = total number of counts (a genes can be present multiple times)

```{r}
p1 <- hist(object@meta.data$nFeature_Spatial)
p2 <- MapFeatures(object = object, features = "nFeature_Spatial", layer = "counts")

p <- p1 | p2
p
```

### Processing

The data colleceted by the array and sequenced after library sequencing are proned to
technical errors. This raw data is called "counts" in a Seurat or semla object.
We need to correct for difference in gene sequence (ScaleData), and in spot differences (NormilizeData).

We checked before by printing the outlook of the object, that we have only the raw counts from after the
spaceranger run.

```{r}
object
```

Now, we will need to multiple steps:

1. Normalize the counts (place holder for the explanation: correct for the gene variance per spot)
2. Scale the data (place holder for explanation why we do that. We use these scale data when further dimensionality reduction with PCA)

```{r}
object <- NormalizeData(object = object)
object <- ScaleData(object = object)
```

We see the difference already by plotting the same gene using the count or the data layer:

```{r}
p1 <- MapFeatures(
  object = object,
  slot = "counts", # Here we use the non-normalized & unscaled layer
  features = "MYH6",
  pt_size = 3,
  image_use = "raw"
)

p2 <- MapFeatures(
  object = object,
  slot = "data", # Here we use the normalized and sclaed layer
  features = "MYH6",
  pt_size = 3,
  image_use = "raw"
)

p <- p1 | p2
p
```

3. Restrict the gene space by only the top variable genes (place holder formula from FindVariableFeatures)
4. Perform Principal Component Analysis (PCA) using the variable genes

```{r}
object <- FindVariableFeatures(object = object, nfeatures = 3000)
object <- RunPCA(object)
```

## Now we will do the same but with 3 sections

### Load data & create infoTable from metadata + spaceranger outs

```{r}
data_root_directory <- file.path("./data/spaceranger/*")
metadata <- read.csv("./data/metadata.csv")

samples <- Sys.glob(paths = file.path(data_root_directory, "filtered_feature_bc_matrix.h5"))
imgs <- Sys.glob(paths = file.path(data_root_directory, "spatial", "tissue_hires_image.png"))
spotfiles <- Sys.glob(paths = file.path(data_root_directory, "spatial", "tissue_positions_list.csv"))
json <- Sys.glob(paths = file.path(data_root_directory, "spatial", "scalefactors_json.json"))
```

```{r}
infoTable <- tibble(samples,
                    imgs,
                    spotfiles,
                    json,
                    slide_id = list.files(path = "./data/spaceranger/"),
                    sample_id = list.files(path = "./data/spaceranger/"),
                    organ = metadata$organ,
                    organism = metadata$organism,
                    age = metadata$pcw_age,
                    sex = metadata$sex)

rm(data_root_directory, samples, imgs, spotfiles, json)
```

### Create semla object & Load images onto object

```{r}
object_3sect <- ReadVisiumData(infoTable)
object_3sect <- LoadImages(object_3sect, image_height = 1931)
object_3sect

rm(infoTable)
gc()
```

### Plot H&E before transformation

```{r}
#| eval: false

ImagePlot(object_3sect)
```

```{r}
MapFeatures(
  object = object_3sect,
  slot = "counts",
  features = "MYH6",
  pt_size = 3,
  image_use = "raw"
)
```

```{r}

```
