---
title: "Step-by-step discovery of Visium data analysis using embryonic heart sections."
author: Raphaël Mauron
date: today
format: html
execute:
  echo: true
editor_options:
  chunk_output_type: console

---

Intro, technology, polyA

## Set up

### Set project library

When working in R, there is multiple ways to set up your working environment. For the 
sake of that workshop, you are provided with a fully working environment that contains
the code, the data, and the R packages already installed. Below we are just specifying
R where to look for the installed pacakges.

```{r}
#| label: setup
#| include: false

.libPaths("/home/raphael.mauron/miniconda3/envs/VisiumHD/lib/R/library") # set project library

.libPaths() # check project library

lapply(.libPaths()[1], list.files) # check library
```


### Load library

R packages might be installed, but for us to be able to use them later in the analysis,
we have to attach them to our session. In R, we do it by using the `library()` function.
This slightly more advanced way below, allow to call the function `library()` in a loop.
You could also do:
`
library(semla)
library(Seurat)
etc.
`
This works equally well.

```{r}
#| label: load packages

package.list <- c("semla", "Seurat", "viridis", "dplyr", "ggplot2", "gtools", "patchwork",
                  "RColorBrewer", "harmony", "tidyverse")
invisible(lapply(package.list, function(element) {
  library(element, character.only = TRUE)
}))
rm(package.list)
```


### Set seed for project

Sometimes, some code and functions rely on some random drawing of numbers.
For instance, if I tell R to sample 100 random cells from my single-cell RNA-seq data,
everytime I run the function, I will have different randomly picked cells. It does what
we ask, but for the sake of reproducibility, we want to have always the same cells picked
when we run the same code the next day. To do so, we can specify a seed by using `set.seed()`.
Some functions have that option in the function's parameter (worth checking because it might
not be used by default), but otherwise you can use `set.seed()` with any number you want 
**in the code chunk where you need it**.

```{r}
set.seed(1)
```


## Loading my first Visium section
Our session is now fully ready to start. We have the packages loaded, we are ready to load
some data.

To analyse a [Visium](https://www.10xgenomics.com/platforms/visium/product-family?&utm_medium=search&src=search&utm_source=google&lss=google&utm_term=10x%20visium&useroffertype=website-page&utm_content=website-page&utm_campaign=701KW0000019A9zYAE&cnm=701KW0000019A9zYAE&cid=701KW0000019A9zYAE&usercampaignid=701KW0000019A9zYAE&gad_source=1) section, we need to load the data.
[10x Genomics](https://www.10xgenomics.com/) provides a software ([spaceranger](https://www.10xgenomics.com/support/software/space-ranger/latest)) that takes the raw sequences and the microscopy images, 
and generates a `count matrix` (gene counts per bacrode), and several files that help us align 
the microscopy image to the count matrix.


### How do we load this count matrix in R?
In R, developpers have developped convenience packages to effortlessly load the `spaceranger`
outputs, store them into an object class, and provide multiple functions to analyse, plot and else.

In the single-cell RNA seq world, such a convenience package is `Seurat`, and is very well established
and maintained by a large team. They are at the origin of the object class that we call `Seurat object`.

When Spatial Transcriptomics appeared, they were no such toolkit, so other labs put effort into creating
such a package. Today we will rely on the [`semla`](https://github.com/spatial-research/semla) package
developped in our lab ([Lundeberg](https://www.spatialresearch.org/)).
`semla`, under the hood, uses the `Seurat object` class, but added a way to register the spatial information
that single-cell data do not have (image, coordinates, etc.).

In the next steps, most functions we will use come from either `semla` or `Seurat`.


### Load data & create infoTable from metadata + spaceranger outs
Let's start!

Below we create a table that provides the path so R (and specific packages e.g., *semla*)
knows where the `spaceranger outputs` are. We have here:

1. `filtered_feature_bc_matrix.h5` : the count data
2. `tissue_hires_image.png` : an H&E image that can be aligned to the counts
3. `tissue_positions_list.csv`: a table that contains the info to align the pixel of the image and the Visium spots
4. `scalefactors_json.json`: a ratio value that help us visualize the image and the spots in the right scale.


```{r}
data_root_directory <- file.path("./data/spaceranger/V10F24-105_C1")

samples <- Sys.glob(paths = file.path(data_root_directory, "filtered_feature_bc_matrix.h5"))
imgs <- Sys.glob(paths = file.path(data_root_directory, "spatial", "tissue_hires_image.png"))
spotfiles <- Sys.glob(paths = file.path(data_root_directory, "spatial", "tissue_positions_list.csv"))
json <- Sys.glob(paths = file.path(data_root_directory, "spatial", "scalefactors_json.json"))
```


We create a table containing all these paths

```{r}
infoTable <- tibble(samples,
                    imgs,
                    spotfiles,
                    json)
```


We can inspect the table:

```{r}
infoTable
```


We can now use `ReadVisiumData()` function from *semla* package to create an R-object with all the data we linked with their paths in the `infoTable`. *semla* uses the data structure from the well-known *Seurat* package, originally developed to facilitate the analysis of *single-cell RNA-seq* data. *semla* added, on top of this architecture, all the spatial information (image, coordinates, etc.) that are required for us to know where are the spots and to be able to later use this information in analyses and visualization.

```{r}
object <- ReadVisiumData(infoTable)
object <- LoadImages(object) #we specify semla to load the image in the object
```


Let's inspect what we have so far.

```{r}
object
```


### Question:

1. What is the dimension of the data?
2. How many spots does that section contain?
3. How many genes does that section contain?
4. Does it make sense?

`Hint`: look at the printed object!


At this stage, we can already use and visualize the data, for instance plotting a gene:

```{r}
MapFeatures(
  object = object,
  slot = "counts",
  features = "MYH6", # <-- place to enter a gene name
  pt_size = 3,
  image_use = "raw"
)
```




### Cleaning up a bit...

We see that there are around the double unique genes compared to the number of human protein coding genes.
Since our data is coming from a poly-A capture technology, every transcripts in the sample 
that binds to a poly-T will be sequenced and counted. Among the extra things, there are some transcripts that we won't want to keep:

- Hemoglobin genes (raison)
- Mitochondrial genes (raison)
- ribosomal genes (raison)
- MALAT1 gene (raison)

In *R*, we can use simple `regular expression` or called `regex`(text patterns that match) to retrieve these genes:

```{r}
# find al the genes that correspond to mt, hb, rp and malat1
undesirable_genes <- grep(pattern = c("^MT-|^HB|^RP|MALAT1"), x = rownames(object), value = T)

# remove two genes that correspond to our search but that we want to keep (growth factor & transcription factor)
undesirable_genes <- setdiff(undesirable_genes, c("HBEGF", "HBP1"))

# check the list of undesirable genes
undesirable_genes
```


We can get a list of all the other genes (the ones we want to keep), and see the difference compared to the original object.

```{r}
genes_to_keep <- setdiff(rownames(object), undesirable_genes)
length(rownames(object)) - length(genes_to_keep)
```

### Question:

1. How many genes is there already that we want to filter out?


Now we can proceed with a first filtering. We want to remove the undesirable genes, meaning we want to keep only all the other genes.
We use the function `SubsetSTData()` from the *semla* package, which ensure that the spatial information is also modified correctly.

```{r}
object <- SubsetSTData(object, features = genes_to_keep)
object
```

Note that when using `SubsetSTData()` we assigned the result of the function to the same original
name `object`. That action overwrites the previous state of our object. In case you want to keep
the unfiltered and filtered together, should should assign the result of the function to a different
variable name.

### More systematic filtering

The extra transcripts we sequence and gather in our object do not limit itself to the 4 categories we succesfully removed above. Databases allow us to annotate the remaining genes we sequenced into some categories.
Here, we will use the  *GRCh38-2020-A genome reference* which let us figure out the main categories our biomolecules belongs too.

The complete *GRCh38-2020-A genome reference* is quite large so here we provide the single file of that reference that you can load in the next chunk. 

*You do not need to run the next chunk*
This chunk commented out is for you to see what has been done from the complete reference to get the 
file we will use next.

```{r}
#| eval: false

# package that help us load large files
library(rtracklayer)

# import the 
gtf <- import("/srv/home/10x_references/refdata-gex-GRCh38-2020-A/genes/genes.gtf")

unique(gtf$type)
# [1] gene           transcript     exon           CDS            start_codon   
# [6] stop_codon     UTR            Selenocysteine

annot <- gtf[gtf$type %in% c("gene")] |>
  mcols() |>
  as.data.frame() |>
  dplyr::select(gene_id, gene_name, gene_type) |>
  dplyr::distinct()

saveRDS(annot, "./data/GRCh38-2020-A_gene_biotypes.rds")
```


Here we load the lighter version of the reference genome:

```{r}
annot <- readRDS("./data/GRCh38-2020-A_gene_biotypes.rds")
head(annot)
unique(annot$gene_type)
```


Here we will use the information of the gene types to keep only the protein coding genes.
This step is project dependent, in other cases you might want to also keep the long-non-coding RNA or any types of immunoglobuline transcripts.

```{r}
# subset the reference genes to only the protein coding genes
annot_protein_coding <- subset(annot, gene_type == "protein_coding")

# fetch the gene names from our object that are annotated as protein coding genes in the reference genome
genes_to_keep_2 <- intersect(rownames(object), annot_protein_coding$gene_name)

# use the same subset idea as before
object <- SubsetSTData(object, features = genes_to_keep_2)
object
```


We can now see that our object size significantly decreased.

### Question:

1. What is the dimension of the data now?
2. How many genes are we going to work with now?


### Some QC plots

We now cleaned the genes that we are not interested in.
We can now analyse the distributions of these genes with the commonly used histograms, but also spatially.

```{r}
# check within the object more in details, we see the metadata
head(object@meta.data)
```

`nFeature_Spatial` = number of unique genes
`nCount_Spatial` = total number of counts (a genes can be present multiple times)

```{r}
p1 <- hist(object@meta.data$nFeature_Spatial)
p2 <- MapFeatures(object = object, features = "nFeature_Spatial", layer = "counts")

p <- p1 | p2
p
```


### Processing

The data colleceted by the array and sequenced after library sequencing are proned to
biological and technical variations. 
This raw data is called "counts" in a `Seurat` object.
We need to correct for differences in gene sequences (ScaleData), and in spot differences (NormilizeData) so that the expression is comparable across sample, spots and genes.

We checked before by printing the outlook of the object, that we have only the raw counts from after the `spaceranger` run.

Now, we will need to multiple steps:


#### 1. Normalization

Normalize the counts (the amount of RNA and UMIs differ between spots (cells in case of single-cell RNA-seq)). The normalization is done per spot and is a log1p of a per spot scaled count.



```{r}
object_norm <- NormalizeData(object = object)
```


```{r}
p1 <- hist(object, n=50) # distribution of the total number of molecule/spot before normalization
p2 <- hist(object_norm, n=50) # distribution of the total number of molecule/spot after normalization

p3 <- MapFeatures(
  object = object,
  slot = "counts", # Here we use the non-normalized & unscaled layer
  features = "nCount_Spatial", #total number of molecules per spot
  pt_size = 3,
  image_use = "raw"
)

p4 <- MapFeatures(
  object = object_norm,
  slot = "data", # Here we use the normalized and sclaed layer
  features = "nCount_Spatial", #total number of molecules per spot
  pt_size = 3,
  image_use = "raw"
)

p <- (p1 | p3) / (p2 | p4)
p
```


#### 2. Find variable features

Restrict the gene space by only the top variable genes (place holder formula from FindVariableFeatures)

Excluding invariable genes that do not contribute
informative/interesting information
– Improved signal -to- noise ratio
– Reduced computational requirements



#### 3. Scaling

The scaling correspond to a Z-score transformation which is performed on the normalized count (after normalization step), and is applied per gene (and not per spot).

Formula:
Z = (x - mu) / sigma

x: normalized count
mu: mean expression of the gene we normalize
sigma: standard deviation of the gene we normalize

Scaled data have a mean of 0 and a variance of 1. That means that some genes have positive values and some are negative. 
This scaled data is used mainly when we want to further reduce dimension using PCA.

```{r}
object_scaled <- ScaleData(object = object_norm)
```

We see the difference already by plotting the same gene using the count or the data slot:

```{r}

```


#### 4. Principal Component Analysis

Perform Principal Component Analysis (PCA) using the variable genes

```{r}
object <- FindVariableFeatures(object = object, nfeatures = 3000)
object <- RunPCA(object)
```

This 4 steps allowed us to correct for sequencing depths difference in genes and in spots.
We found the genes that have the most amount of variance in the data (the genes that do not
change in space in our section are not the most interesting), and performed PCA.
PCA tries to get the top variable genes, and tries to 




## Now we will do the same but with 3 sections

### Load data & create infoTable from metadata + spaceranger outs

```{r}
data_root_directory <- file.path("./data/spaceranger/*")
metadata <- read.csv("./data/metadata.csv")

samples <- Sys.glob(paths = file.path(data_root_directory, "filtered_feature_bc_matrix.h5"))
imgs <- Sys.glob(paths = file.path(data_root_directory, "spatial", "tissue_hires_image.png"))
spotfiles <- Sys.glob(paths = file.path(data_root_directory, "spatial", "tissue_positions_list.csv"))
json <- Sys.glob(paths = file.path(data_root_directory, "spatial", "scalefactors_json.json"))
```

```{r}
infoTable <- tibble(samples,
                    imgs,
                    spotfiles,
                    json,
                    slide_id = list.files(path = "./data/spaceranger/"),
                    sample_id = list.files(path = "./data/spaceranger/"),
                    organ = metadata$organ,
                    organism = metadata$organism,
                    age = metadata$pcw_age,
                    sex = metadata$sex)

rm(data_root_directory, samples, imgs, spotfiles, json)
```

### Create semla object & Load images onto object

```{r}
object_3sect <- ReadVisiumData(infoTable)
object_3sect <- LoadImages(object_3sect, image_height = 1931)
object_3sect

rm(infoTable)
gc()
```

### Plot H&E before transformation

```{r}
#| eval: false

ImagePlot(object_3sect)
```

```{r}
MapFeatures(
  object = object_3sect,
  slot = "counts",
  features = "MYH6",
  pt_size = 3,
  image_use = "raw"
)
```

```{r}

```
